---
title: "exploratory data analysis"
author: "Data Punk"
date: "2025-05-14"
output: html_document
---

License Information

MIT License

Copyright (c) 2024 Data Punk Hockey

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all
copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
SOFTWARE.


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Load core libraries. 

```{r}

library(tidyverse)
library(ggplot2)
library(readr)
library(corrplot)

```

Load data from Hockey Reference into an R data frame and re-factor PLAYOFF data. 

```{r}

nhl_team_stats_df <- read_csv("all_stats_data.csv")

nhl_team_stats_df$PLAYOFF <- factor(nhl_team_stats_df$PLAYOFF, labels = c("No", "Yes"))

```

Plot Average Points Percentage by Playoff Status. 

```{r}

ggplot(nhl_team_stats_df, aes(x = PLAYOFF, y = PTS_PCT, fill = PLAYOFF)) +
  geom_boxplot(width = 0.5, outlier.shape = 21, outlier.fill = "white", outlier.size = 2) +
  scale_fill_manual(values = c("No" = "#FF9999", "Yes" = "#66CC99")) +
  labs(
    title = "PTS PCT by Playoff Status",
    x = "Playoff Team?",
    y = "PTS PCT"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    axis.text = element_text(size = 12),
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )


```
Plot Average Goals For Per Game vs. Goals Against Per Game.

```{r}

ggplot(nhl_team_stats_df, aes(x = GFPG, y = GAPG, color = PLAYOFF)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_manual(values = c("No" = "#FF6666", "Yes" = "#339966")) +
  labs(
    title = "Goals For vs. Goals Against per Game",
    subtitle = "Playoff Teams vs. Non-Playoff Teams",
    x = "Goals For Per Game (GFPG)",
    y = "Goals Against Per Game (GAPG)",
    color = "Playoff Team"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, margin = margin(b = 10)),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    axis.text = element_text(size = 12),
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_blank(),
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )


```
Plot Power Play and Penalty Kill Percentage by Playoff Status. 

```{r}

nhl_team_stats_df %>%
  pivot_longer(cols = c(PP_PCT, PK_PCT), names_to = "Special_Teams", values_to = "Percentage") %>%
  ggplot(aes(x = PLAYOFF, y = Percentage, fill = PLAYOFF)) +
  geom_boxplot() +
  facet_wrap(~Special_Teams, scales = "free_y") +
  labs(title = "Special Teams Performance by Playoff Status", x = "Made Playoffs?", y = "Percentage") +
  theme_minimal()

```

Create Correlation matrix for numeric performance metrics. 

```{r}


library(corrplot)
library(RColorBrewer)

cor_df <- nhl_team_stats_df %>%
  select(GFPG, GAPG, PP_PCT, PK_PCT, S_PCT, SV_PCT, PDO, CF_PCT, XGF, XGA, SCF_PCT, HDF_PCT)

cor_matrix <- cor(cor_df, use = "complete.obs")

corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  col = colorRampPalette(brewer.pal(n = 8, name = "RdYlBu"))(100),
  tl.col = "black",
  tl.cex = 1,
  cl.cex = 0.8,
  number.cex = 0.8,
  addCoef.col = "black",
  diag = FALSE,
  title = "Correlation of Key Performance Metrics",
  mar = c(0, 0, 2, 0)  # Adjusts margin for better title display
)


```

Create Logistic regression to predict playoff probability.

```{r}

model <- glm(PLAYOFF ~ GFPG + PDO + SV_PCT, 
             data = nhl_team_stats_df, family = "binomial")
summary(model)

```

Plot Predicted probabilities of making playoffs

```{r}

nhl_team_stats_df$playoff_prob <- predict(model, type = "response")

ggplot(nhl_team_stats_df, aes(x = playoff_prob, fill = PLAYOFF)) +
  geom_histogram(binwidth = 0.05, alpha = 0.6, position = "identity", color = "white") +
  labs(title = "Predicted Playoff Probability vs Actual", x = "Predicted Probability", y = "Count") +
  theme_minimal()

```


