---
title: "predictive_model"
author: "Data Punk"
date: "2025-05-17"
output: html_document
---

License Information

MIT License

Copyright (c) 2024 Data Punk Hockey

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all
copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
SOFTWARE.

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

 
# Load libraries for the script.

```{r}

library(tidyverse)
library(dplyr)
library(rvest)
library(xml2)

```

# Scrape the data from Hockey-Reference.com. 

```{r}

url <- "https://www.hockey-reference.com/playoffs/NHL_2025.html"

team_statistics <- url %>% 
  read_html() %>% 
  html_nodes(xpath = '//*[@id="teams"]') %>% 
  html_table()

playoff_statistics <- team_statistics[[1]]

playoff_pwp_df <- as.data.frame(playoff_statistics)

playoff_pwp_df <- playoff_pwp_df %>% 
  filter(Team != "League Average")

print(playoff_pwp_df)

colnames(playoff_pwp_df) <- c("RANK", "TEAM", "GP", "W", "L", "T", "OW", "OL",
                              "WL_PCT", "GF", "GA", "DIFF")


```



```{r}

team_abbr_lookup <- data.frame(
  TEAM = c("Carolina Hurricanes", "Edmonton Oilers", "Florida Panthers", 
           "Dallas Stars", "Toronto Maple Leafs", "Winnipeg Jets",
           "Washington Capitals", "Vegas Golden Knights",
           "Colorado Avalanche", "St. Louis Blues", "Ottawa Senators", 
           "Minnesota Wild", "Los Angeles Kings", "New Jersey Devils", 
           "Montreal Canadiens", "Tampa Bay Lightning"),
  TEAM_ABB = c("CAR", "EDM", "FLA", "DAL", "TOR", "WPG", "WSH", "VGK", 
               "COL", "STL", "OTT", "MIN", "LAK", "NJD", "MTL", "TBL"),
  stringsAsFactors = FALSE
)

playoff_pwp_df <- playoff_pwp_df %>%
  left_join(team_abbr_lookup, by = "TEAM")

playoff_pwp_df$SEASON = "2024-2025"

playoff_pwp_df$PWP = (playoff_pwp_df$GF^2)/(playoff_pwp_df$GF^2+playoff_pwp_df$GA^2)

playoff_pwp_df <- playoff_pwp_df %>% 
  select(SEASON, RANK, TEAM, TEAM_ABB, GP, W, L, T, OW, OL, WL_PCT, GF, GA,
         DIFF, PWP)

print(playoff_pwp_df)

write.csv(playoff_pwp_df, "current_playoff_pwp_df.csv", row.names = FALSE)

```


# Optionally load the data manually and filter the data for what you need.

```{r}

# playoff_pwp_data <- read.csv("playoff_pwp.csv")

team_summary <- playoff_pwp_df %>% 
  select(SEASON, TEAM_ABB, GF, GA, PWP)

print(team_summary)

write.csv(team_summary, "playoff_team_summary.csv", row.names = FALSE)

```

# Use the PWP to predict the likelihood of a team winning the Cup. 

```{r}

predict_stanley_cup_chance <- function(team, season) {
  df_prediction <- team_summary %>% 
    filter(TEAM_ABB == team, SEASON == season)
  
  if (nrow(df_prediction) == 0) {
    return("No data available for this team.")
  }
  
  pwp_score <- df_prediction$PWP
  message <- ifelse(pwp_score > 0.6, 
                    "High likelihood of winning the Cup.", 
                    ifelse(pwp_score > 0.5, 
                           "Moderate chance of winning the Cup.", 
                           "Low chance of winning the Cup."))
  
  cat("\n=== Stanley Cup Prediction ===\n")
  cat("Team: ", team, "\nSeason: ", season, "\n")
  cat("PWP:", round(pwp_score, 3), "\n")
  cat("Predicted Outcome: ", message, "\n")
}

```



```{r}

predict_stanley_cup_chance("CAR", "2024-2025")

```



```{r}

predict_win_percentage_vs_field <- function(team, season) {
  playoff_teams <- team_summary %>%
    filter(SEASON == season)

  if (!(team %in% playoff_teams$TEAM)) {
    return("Team not found.")
  }

  team_pwp <- playoff_teams %>% filter(TEAM_ABB == team) %>% pull(PWP)
  total_pwp <- sum(playoff_teams$PWP, na.rm = TRUE)
  
  win_pct <- team_pwp / total_pwp * 100
  
  cat("\n=== Championship Probability Based on PWP ===\n")
  cat("Team: ", team, "\n")
  cat("Season: ", season, "\n")
  cat("PWP: ", round(team_pwp, 3), "\n")
  cat("Predicted Chance of Winning Stanley Cup: ", sprintf("%.2f%%", win_pct), "\n")
}

```


```{r}

predict_win_percentage_vs_field("CAR", "2024-2025")

```


```{r}

rank_stanley_cup_chances <- function(season) {
  playoff_teams <- team_summary %>%
    filter(SEASON == season) %>%
    drop_na(PWP) 
  
  total_pwp <- sum(playoff_teams$PWP, na.rm = TRUE)
  
  ranked_df <- playoff_teams %>%
    mutate(
      Win_Percentage = (PWP / total_pwp) * 100
    ) %>%
    arrange(desc(Win_Percentage)) %>%
    select(TEAM_ABB, PWP, Win_Percentage)
  
  return(ranked_df)
}

```


```{r}

ranked_chances_2024 <- rank_stanley_cup_chances("2024-2025")
print(ranked_chances_2024)

```



```{r}

rank_stanley_cup_chances <- function(season, playoff_teams_vec) {
  playoff_teams <- team_summary %>%
    filter(SEASON == season, TEAM_ABB %in% playoff_teams_vec) %>%
    drop_na(PWP)
  
  total_pwp <- sum(playoff_teams$PWP, na.rm = TRUE)
  
  ranked_df <- playoff_teams %>%
    mutate(
      Win_Percentage = (PWP / total_pwp) * 100
    ) %>%
    arrange(desc(Win_Percentage)) %>%
    select(TEAM_ABB, PWP, Win_Percentage)
  
  return(ranked_df)
}

```



```{r}

remaining_teams <- c("CAR", "FLA", "DAL", "EDM")

ranked_playoff_chances <- rank_stanley_cup_chances("2024-2025", remaining_teams)

print(ranked_playoff_chances)

```




```{r}

head_to_head_prediction <- function(team1, team2, season) {
  df1 <- team_summary %>% filter(TEAM_ABB == team1, SEASON == season)
  df2 <- team_summary %>% filter(TEAM_ABB == team2, SEASON == season)
  
  if (nrow(df1) == 0 || nrow(df2) == 0) {
    return("One or both teams not found.")
  }
  
  pwp1 <- df1$PWP
  pwp2 <- df2$PWP
  
  prob_team1_win <- pwp1 / (pwp1 + pwp2)
  prob_team2_win <- pwp2 / (pwp1 + pwp2)
  
  cat("\n=== Head-to-Head Prediction ===\n")
  cat("Playoffs:", season, "\n")
  cat(team1, "Win Probability:", sprintf("%.1f%%", 100 * prob_team1_win), "\n")
  cat(team2, "Win Probability:", sprintf("%.1f%%", 100 * prob_team2_win), "\n")
}

```


```{r}

head_to_head_prediction("CAR", "FLA", "2024-2025")

```




