---
title: "cluster_analysis"
author: "Data Punk"
date: "2024-06-29"
output: html_document
---

License Information

MIT License

Copyright (c) 2023 Data Punk Hockey

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all
copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
SOFTWARE.

# Libraries needed for cluster analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(tidyverse)
library(cluster)
# library(dendextend)
library(knitr)
library(kableExtra)

```

# Load and filter data.

```{r}

player_data <- read_csv('multi_season_player_data.csv')

subset_player_data <- player_data %>% 
  filter(POS == "C") %>% 
  filter(GP > 20) %>% 
  filter(SEASON == "2023_2024")

data_sniper <- subset_player_data %>% 
  select(NAME, TEAM, GP, SHOT_PCT, PTS) %>% 
  drop_na()
  

str(data_sniper)

```

# Determine the number of clusters for k-means analysis.

```{r}

set.seed(42)
sse <- vector()
k_range <- 1:10

for (k in k_range) {
  kmeans_model <- kmeans(data_sniper$SHOT_PCT, centers = k, nstart = 25)
  sse[k] <- sum(kmeans_model$tot.withinss)
}

ggplot(data.frame(k = k_range, sse = sse), aes(x = k, y = sse)) +
  geom_line(color = "lightblue", size = 1) +
  geom_point(size = 2, color = "blue") +
    labs(
    title = 'Elbow Method for Optimal k',
    x = 'Number of Clusters (k)',
    y = 'Sum of Squared Distances (SSE)'
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 8),
    panel.grid.major = element_line(color = "gray88"),
    panel.grid.minor = element_line(color = "gray95")
  ) +
  scale_x_continuous(breaks = k_range)

```

# Run k-means analysis (and recode clusters to be more intuitive).

```{r}

set.seed(42)
kmeans_final <- kmeans(data_sniper$SHOT_PCT, centers = 5, nstart = 25)

data_sniper_cluster <- data_sniper %>% 
  mutate(Cluster = as.factor(kmeans_final$cluster)) %>% 
  mutate(Cluster = recode(Cluster,
                          `3` = "L5",
                          `1` = "L4",
                          `2` = "L3",
                          `4` = "L2",
                          `5` = "L1"))

ggplot(data_sniper_cluster, aes(x = NAME, y = SHOT_PCT, color = Cluster)) +
  geom_point() +
  labs(title = 'K-Means Clustering of NHL Snipers', x = 'Player', y = 'Shot Percentage') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, hjust = 1), axis.text.y = element_text(angle = 0, hjust = 1, size = 4)) +
  coord_flip()


write.csv(data_sniper_cluster, "cluster_of_snipers.csv", row.names = FALSE)

```
# Create bar chart of cluster sizes.

```{r}

count_of_clusters_df <- data_sniper_cluster %>% 
  group_by(Cluster) %>% 
  summarise(TOTAL_COUNT = n(), AVG_SHOT_PCT = round(mean(SHOT_PCT),2)) %>% 
  arrange(desc(AVG_SHOT_PCT))

total_count <- sum(count_of_clusters_df$TOTAL_COUNT)

count_of_clusters_df <- count_of_clusters_df %>%
  mutate(Percentage = TOTAL_COUNT / total_count * 100)

count_of_clusters_df$Percentage <- round(count_of_clusters_df$Percentage, 2)

count_of_clusters_df <- count_of_clusters_df %>%
  mutate(Cluster = fct_reorder(Cluster, TOTAL_COUNT))

ggplot(count_of_clusters_df, aes(x = Cluster, y = TOTAL_COUNT, fill = Cluster)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = paste0(TOTAL_COUNT, " (", round(Percentage, 1), "%)")), vjust = -0.5, size = 4) +
  labs(title = "Count of Snipers by Cluster", x = "Cluster", y = "Total Count") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 8),
    panel.grid.major = element_line(color = "gray88"),
    panel.grid.minor = element_line(color = "gray95"),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  scale_fill_brewer(palette = "Set3") +
  ylim(0, max(count_of_clusters_df$TOTAL_COUNT) * 1.1)

```

# Focus on top 10 players in top cluster.

```{r}

sub_final_cluster <- data_sniper_cluster %>% 
  filter(Cluster == "L5") %>% 
  arrange(desc(PTS))

sub_final_cluster

top_10_records <- sub_final_cluster %>%
  top_n(10, SHOT_PCT) %>% 
  arrange(desc(SHOT_PCT))

top_10_records

top_10_records %>%
  kable("html") %>%
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(1, bold = T) %>%
  row_spec(0, bold = T, background = "#D3D3D3")


```







