---
title: "Predicting Wins - Logistic Regression"
author: "Data Punk"
date: "2024-07-13"
output: html_document
---

MIT License

Copyright (c) 2024 Data Punk Hockey

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all
copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
SOFTWARE.


# Load the libraries for the logistic regression analysis.

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
library(caret)
library(glmnet)
library(pROC)
library(tidyverse)
library(reshape2)
library(knitr)
library(kableExtra)
library(broom)

```

# Load, clean and transform the data.

```{r}
final_log_reg_df <- read.csv("final_log_regression_game_data.csv")

sub_final_log_reg_df <- final_log_reg_df %>% 
  select(Date, Team, Win, GF, GA, SF_PCT, xGF, xGA, xGF_PCT, CF_PCT, FF_PCT, SCF_PCT, HDSF_PCT,
         HDGF_PCT)

colnames(sub_final_log_reg_df) <- c("DATE", "TEAM", "WIN", "GF", "GA", 
                                    "SF_PCT", "X_GF", "X_GA", "XGF_PCT", 
                                    "CF_PCT", "FF_PCT", "SCF_PCT", "HDSF_PCT",
                                    "HDGF_PCT")

sub_final_log_reg_df <- sub_final_log_reg_df %>%
  mutate(HDGF_PCT = ifelse(HDGF_PCT == "-", 0, as.numeric(HDGF_PCT)))

na.omit(sub_final_log_reg_df)

sub_final_log_reg_df

write.csv(sub_final_log_reg_df, "subset_log_regression_game_data.csv", row.names = FALSE)

```

# Create the training and test data.

```{r}

set.seed(123)

train_index <- createDataPartition(sub_final_log_reg_df$WIN, p = .8, list = FALSE)

data_train_df <- sub_final_log_reg_df[train_index, ]
data_test_df <- sub_final_log_reg_df[-train_index, ]

```

#  Create the model.

```{r}

model <- glm(WIN ~ SF_PCT + XGF_PCT + CF_PCT + FF_PCT + HDGF_PCT, 
             data = data_train_df, family = binomial)

tidy_model <- tidy(model)

tidy_model %>%
  kable(format = "html", caption = "Logistic Regression Model Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) %>%
  column_spec(1, bold = TRUE, color = "white", background = "dodgerblue") %>%
  column_spec(2:5, color = "black", background = "lightgray")

```


# Create the predictions and test actuals versus predictions.

```{r}

data_test_df <- data_test_df[, names(data_train_df)]

predictions <- predict(model, newdata = data_test_df, type = "response")
predicted_classes <- ifelse(predictions > 0.5, 1, 0)


result_summary <- data.frame(
  Actual = data_test_df$WIN,
  Predicted = predicted_classes
)

result_summary

kable(result_summary, format = "html", caption = "Actual vs Predicted Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) %>%
  column_spec(1, bold = TRUE, color = "white", background = "dodgerblue") %>%
  column_spec(2, bold = TRUE, color = "white", background = "orange")

```

# Create the confusion matrix.

```{r}

conf_matrix <- table(Predicted = predicted_classes, Actual = data_test_df$WIN)

conf_matrix_df <- as.data.frame(as.table(conf_matrix))

ggplot(data = conf_matrix_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), vjust = 1) +
  scale_fill_gradient(low = "lightblue", high = "orange") +
  labs(title = "Confusion Matrix", x = "Actual", y = "Predicted") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    axis.text = element_text(size = 10)
  )

```
# Evaluate the model metrics to test specificity and sensitivity.

```{r}

confusion_matrix <- confusionMatrix(factor(predicted_classes), factor(data_test_df$WIN))

performance_metrics <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity", "Pos Pred Value", "Neg Pred Value", "Kappa", "Balanced Accuracy"),
  Value = c(
    confusion_matrix$overall["Accuracy"],
    confusion_matrix$byClass["Sensitivity"],
    confusion_matrix$byClass["Specificity"],
    confusion_matrix$byClass["Pos Pred Value"],
    confusion_matrix$byClass["Neg Pred Value"],
    confusion_matrix$overall["Kappa"],
    confusion_matrix$byClass["Balanced Accuracy"]
  )
)

kable(performance_metrics, format = "html", caption = "Performance Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) %>%
  column_spec(1, bold = TRUE, color = "white", background = "dodgerblue") %>%
  column_spec(2, bold = TRUE, color = "white", background = "orange")

```

# Plot the ROC curve.

```{r}

roc_curve <- roc(data_test_df$WIN, predictions)
auc_value <- auc(roc_curve)

roc_plot <- ggroc(roc_curve, color = "blue", size = 1.2) +
  geom_abline(linetype = "dashed", color = "red") +
  labs(title = "ROC Curve", 
       subtitle = paste("AUC =", round(auc_value, 2)),
       x = "False Positive Rate", 
       y = "True Positive Rate") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 15),
    axis.title.x = element_text(size = 15),
    axis.title.y = element_text(size = 15),
    axis.text = element_text(size = 12),
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_line(color = "gray95")
  ) +
  annotate("text", x = 0.7, y = 0.2, label = paste("AUC =", round(auc_value, 2)), color = "blue", size = 5, hjust = 0)

print(roc_plot)

```


